\section{Asymptotic Complexity Analysis}
\label{sec:asymptotic}

Monads are known for their versatility in representing various semantics,
including side effects. So far, we have made use of our semantics monad for
representing partiality, \ie nontermination and abnormal termination. We may in
the future extend the monad to reason about more effects such as (unsafe) mutable global
variables or I/O. In this section, we instead make use of the monad for verifying a
property different from functional correctness: runtime complexity.

\subsection{Classifying Asymptotic Complexity}

We start with a formalization of multiparametric asymptotic function analysis
based on the ideas and the Coq implementation from~\cite{gueneau}. The main insight of that report is that
we can elegantly formalize the notation of ``going to infinity'' for an
arbitrary number of parameters using the mathematical concept of \emph{filters},
originally from topology.

We refer to the report for a detailed description of
filters. Fortunately for us, the Lean library already includes
a definition of filters on sets. We will only need the filter \lean{at_infty} on
natural numbers, the filter combinator \lean{prod_filter}, which we developed,
and the ``eliminator'' \lean{eventually}.

\begin{minted}{lean}
definition at_infty : filter â„• := ...
notation `[at ` `âˆ]`  := at_infty

definition prod_filter {A B : Type} (Fa : filter A) (Fb : filter B) : filter (A Ã— B) := ...
notation `[at ` `âˆ` ` Ã— ` `âˆ]` := prod_filter [at âˆ] [at âˆ]

definition eventually {A : Type} (P : A â†’ Prop) (F : filter A) : Prop := ...
\end{minted}

A proposition such as \lean{eventually P [at âˆ Ã— âˆ]} then has the intuitive meaning
of holding iff there exists a pair of natural numbers such that \lean{P} holds
for all (componentwise) larger pairs.

We can now formalize the notions of a function (into the natural numbers) being non-strictly and strictly
asymptotically bounded by another function, which directly lead to the usual
notations as classes of functions.

\begin{minted}{lean}
namespace asymptotic
  parameters {A : Type} (F : filter A)
  variables  (f g : A â†’ â„•)

  definition le : Prop := âˆƒ c : â„•, eventually {a | f a â‰¤ c * g a} F
  -- bring `c` to the other side of the inequality
  -- so that we can remain within integer arithmetics
  definition lt : Prop := âˆ€ c : â„•, eventually {a | c * f a â‰¤ g a} F
  definition equiv : Prop := le f g âˆ§ le g f

  definition ub  := {f | le f g}
  definition sub := {f | lt f g}
  definition lb  := {f | le g f}
  definition slb := {f | lt g f}
end asymptotic

notation `ğ“(` g `) ` F := asymptotic.ub F g
notation `ğ“¸(` g `) ` F := asymptotic.sub F g
notation `Î©(` g `) ` F := asymptotic.lb F g
notation `Ï‰(` g `) ` F := asymptotic.slb F g
notation `Î˜(` g `) ` F := asymptotic.equiv F g
\end{minted}

With the notations in place, we can prove familiar lemmas about combining
complexity bounds for arbitrary functions and filters and lemmas about bounds for some
specific functions and filters.

\begin{minted}{lean}
lemma ub_subset_ub : f âˆˆ ğ“(g) F â†’ ğ“(f) F âŠ† ğ“(g) F := ...
lemma add_ub : fâ‚ âˆˆ ğ“(g) F â†’ fâ‚‚ âˆˆ ğ“(g) F â†’ fâ‚ + fâ‚‚ âˆˆ ğ“(g) F := ...
lemma ub_add_left : ğ“(gâ‚‚) F âŠ† f âˆˆ ğ“(gâ‚ + gâ‚‚) F := ...
lemma ub_add_const : fâ‚ âˆˆ ğ“(g) F âˆ© Î©(Î» x, k) F â†’
  fâ‚ + (Î» x, k) âˆˆ ğ“(g) F âˆ© Î©(Î» x, k) F := ..
lemma const_ub_one : (Î» x, k) âˆˆ ğ“(1) F := ...
lemma ub_mul_prod_filter : fâ‚ âˆˆ ğ“(gâ‚) Fâ‚ â†’ fâ‚‚ âˆˆ ğ“(gâ‚‚) Fâ‚‚ â†’
  (Î» p, fâ‚ p.1 * fâ‚‚ p.2) âˆˆ ğ“(Î» p, gâ‚ p.1 * gâ‚‚ p.2) (prod_filter Fâ‚ Fâ‚‚) := ...

lemma log_unbounded {b : â„•} (H : b > 1) : log b âˆˆ Ï‰(1) [at âˆ] := ...
lemma id_unbounded : id âˆˆ Ï‰(1) [at âˆ] := ...
\end{minted}

\subsection{Verifying the Complexity of \texttt{[T]::binary\_search}}

As described in \autoref{sec:sem}, our semantics monad contains a step counter
that is incremented on each function call and loop iteration. Because only a
constant number of instructions can be executed between any such two events for
a given program, the step count of an execution is asymptotically equivalent to the instruction
count, which in turn is usually assumed to be asymptotically equivalent to the running time.

We extend our existing correctness proof of \lean{binary_search} by introducing
a new predicate that tests both the return value and the step count.

\begin{minted}{lean}
inductive sem.terminates_with_in {A : Typeâ‚} (H : A â†’ Prop) (max_cost : â„•) : sem A â†’ Prop :=
mk : Î  {x k}, H x â†’ k â‰¤ max_cost â†’ sem.terminates_with_in H max_cost (some (x, k))
\end{minted}

Because we will only prove asymptotic upper bounds, we also use an upper bound
in the definition in order to simplify reasoning about specific cost functions.
If we wanted to use the predicate with operators other than \lean{ğ“}, we should turn the inequality
into an equality.

This time we analyze the function bottom-up, starting with a single loop iteration, \ie
an execution of \lean{loop_4}. With all dependencies unfolded, we quickly obtain a
constant bound on the step count for everything except the trait call to
\lean{Ord.cmp}, of whose complexity we have absolutely no information. If we wanted
to obtain the textbook bound of $\mathcal{O}(\log n)$ for binary search, we would
have to assume that comparing two elements takes only constant time. That is
certainly not true for all implementations of the trait and such a
restriction would be a shame since for the correctness theorem we did a general proof for any
decidable linear order. Thus we instead introduce a more dynamic upper bound for
the call: the maximum of all execution costs of such comparisons.

\begin{minted}{lean}
-- recall our extension of `Ord` from Section 5.3
structure Ord' [class] (T : Typeâ‚) extends Ord T, decidable_linear_order T :=
(cmp_eq : âˆ€ x y : T, Î£ k, cmp x y = some (ordering x y, k))

definition Ord'.cmp_max_cost {T : Typeâ‚} [Ord' T] (y : T) (xs : list T) :=
-- extracts `k` from the above definition
Max x âˆˆ to_finset xs, sigma.pr1 (cmp_eq x y)
\end{minted}

Now we can prove a specific upper bound of \lean{Ord'.cmp_max_cost needle self +
  15} for the loop body. Finally, we abstract from this explicit cost function
to an asymptotic bound.

\begin{minted}{lean}
lemma loop_4.spec :
  âˆƒ c âˆˆ ğ“(id) [at âˆ],
  âˆ€ self needle s base, sorted le self â†’ is_slice self â†’ loop_4_invar self needle s base â†’
  sem.terminates_with_in
    (loop_4_res self needle s)
    (c (Ord'.cmp_max_cost needle self))
    (loop_4 (closure_5642.mk needle, base, s)) :=
exists.intro (Î» n, n + 15) ...
\end{minted}

This lemma says that the execution cost of \lean{loop_4} is linearly bound by
the maximum comparison cost. In general, we have to separate the \emph{measure}
function that reduces the input data to a natural number (here
\lean{cmp_max_cost}) and the abstract cost function that describes
the asymptotic behavior of the measure result, since we cannot define the latter
on arbitrary domains. The composition of both then gives us the actual upper
bound function.

We also have to make sure to introduce any parameters the
measure depends on only after the existential quantifier. This makes the
definitions slightly more verbose since we cannot use the convenient
\lean{section} mechanisms with them any more.

Going up to the whole loop, we expect its asymptotic running time to be that of the body
multiplied with \lean{logâ‚‚ (length self)}. Formally, we again have to split the
measure function \lean{length} from the asymptotic cost function \lean{logâ‚‚}.

\begin{minted}{lean}
lemma loop_loop_4.spec :
  âˆƒâ‚€ f âˆˆ ğ“(Î»p, logâ‚‚ p.1 * p.2) [at âˆ Ã— âˆ],
  âˆ€ self needle, is_slice self â†’ sorted le self â†’ sem.terminates_with_in
    (binary_search_res self needle)
    (f (length self, Ord'.cmp_max_cost needle self))
    (loop loop_4 (closure_5642.mk needle, 0, self)) := ...
\end{minted}

As in the functional correctness proof, we can show this lemma by well-founded
recursion. However, proving that a loop is asymptotically bounded by an
iteration upper bound multiplied by an upper bound for the body should be a
common occurence, so we have extracted the proof into a general theorem.

\begin{minted}{lean}
theorem loop.terminates_with_in_ub
  {In State Res : Typeâ‚}
  (body : In â†’ State â†’ sem (State + Res))
  (pre : In â†’ State â†’ Prop)
  (p : In â†’ State â†’ State â†’ Prop)
  (q : In â†’ State â†’ Res â†’ Prop)
  (citer aiter : â„• â†’ â„•)
  (miter : State â†’ â„•)
  (cbody abody : â„• â†’ â„•)
  (mbody : In â†’ State â†’ â„•)
  (citer_aiter : citer âˆˆ ğ“(aiter) [at âˆ] âˆ© Î©(1) [at âˆ])
  (cbody_abody : cbody âˆˆ ğ“(abody) [at âˆ] âˆ© Î©(1) [at âˆ])
  (pre_p : âˆ€ args s, pre args s â†’ p args s s)
  (step : âˆ€ args init s, pre args init â†’ p args init s â†’
    sem.terminates_with_in (Î» x, match x with
      | inl s' := p args init s' âˆ§ citer (miter s') < citer (miter s)
      | inr r  := q args init r
      end) (cbody (mbody args init)) (body args s)) :
  âˆƒ f âˆˆ ğ“(Î» p, aiter p.1 * abody p.2) [at âˆ Ã— âˆ], âˆ€ args s, pre args s â†’
    sem.terminates_with_in (q args s) (f (miter s, mbody args s))
      (loop (body args) s) := ...
\end{minted}

This may very well be the most complex theorem of our work, at least by
signature. Going through the explicit parameters from top to bottom, we have the
loop body, the precondition, the invariant (which may depend on both the initial
and current state), the postcondition, the concrete and asymptotic bound and
measure function of the iteration count, and the same for the body. These are
followed by assumptions that the asymptotic bounds are correct, that the
precondition implies the invariant, and that a loop iteration either continues
the loop with the invariant upheld and the concrete iteration count reduced or
breaks the loop while satisfying the postcondition. In the end, the conclusion says
that the loop, measured by the product of the measure functions, is
asymptotically bounded by the product of the asymptotic bounds and terminates
with the postcondition fulfilled, as long as the precondition is satisfied.

When using this theorem to prove the previous lemma, we can transfer the
instantiations of and proofs about the precondition, invariant, and
postcondition directly from the correctness proof, and show the asymptotic
behavior of the body from the lemma \lean{loop_4.spec}. We are left to prove
that the iteration count is asymptotically bounded by \lean{logâ‚‚}. Because this
is the more interesting bound, we will show some more details of the proof.

We choose the concrete bound \lean{Î» n, logâ‚‚ (2 * n) + 1} for the iteration
count and show that it is in \lean{ğ“(logâ‚‚) [at âˆ]}. Because the underlying relation
\lean{asymptotic.le} is transitive, we can make use of Lean's \lean{calc} blocks
for this. By a second transitivity lemma, we can even combine it with the
standard (pointwise) function inequality operator.

\begin{minted}{lean}
local infix `â‰¼`:25 := asymptotic.le [at âˆ]
...

calc (Î» n, logâ‚‚ (2 * n) + 1)
    â‰¤ (Î» n, logâ‚‚ n + 2) : ...
... â‰¼ logâ‚‚ : add_ub asymptotic.le.refl (
      calc (Î» n, 2) â‰¼ (Î» n, 1) : const_ub_one
                ... â‰¼ logâ‚‚     : asymptotic.le_of_lt (log_unbounded ...))
\end{minted}

Finally, we show that the concrete iteration count is strictly decreasing. This
follows from the premises \lean{length s' â‰¤ length s / 2} and \lean{length s â‰  0} of \lean{loop_4_step} from the correctness proof, together with the fact
that \lean{logâ‚‚} is monotone.

\begin{minted}{lean}
calc logâ‚‚ (2 * length s') + 1
    â‰¤ logâ‚‚ (length s) + 1     : ...`length s' â‰¤ length s / 2`...
... = logâ‚‚ (2 * length s)     : ...`length s â‰  0`...
... < logâ‚‚ (2 * length s) + 1 : ...
\end{minted}

Going up from the loop to \lean{binary_search}, there is just a single call, leaving
the final asymptotic complexity unchanged.

\begin{minted}{lean}
theorem binary_search.spec :
  âˆƒ f âˆˆ ğ“(Î» p, logâ‚‚ p.1 * p.2) [at âˆ Ã— âˆ],
  âˆ€ (self : slice T) (needle : T), is_slice self â†’ sorted le self â†’ sem.terminates_with_in
    (binary_search_res self needle)
    (f (length self, Ord'.cmp_max_cost needle self))
    (binary_search self needle) := ...
\end{minted}

Thus, in the most general way, the runtime of \lean{binary_search} is
asymptotically bounded by the logarithm of the input slice's length multiplied
with the maximum cost of comparing the needle with any element in the slice.